---
title: "Seminar: Statistics for Data Science: Group C"
subtitle: "Personal Agency vs. Environmental Determinism in Academic Success"
author: "Dmytro Tupkalenko (89252101), Natalia Tashkova (89252098), Yllkë Jashari (89252111)"
date: "May 2026"
date-format: "MMMM YYYY"
format: 
  html:
    toc: true
    toc_float: yes
    toc_depth: 3
    embed-resources: true
    theme: "lux"
    css: styles.css
    code-fold: true
    code-summary: "Show code"
execute: 
  warning: false
  message: false
  echo: true
  error: true
  cache: true
knitr:
  opts_chunk:
    dpi: 300
editor: 
  markdown: 
    wrap: 72
---

## Setup

```{r}
#| label: setup

library(tidyverse)
library(naniar)  
library(Hmisc)     
library(gt)
library(patchwork)
library(viridis)
library(pheatmap)
library(RColorBrewer)
library(broom)

VIR_PAL = "B"
theme_set(theme_minimal(base_size = 11))
setwd("./")

```

## Research Question & Statistical Analysis Plan

### Aim

General Formulation --- Personal Agency vs. Environmental Determinism:
To what extent do student-driven behavioral choices predict academic
success in Portuguese and Mathematics compared to structural and
environmental constraints?

Concrete Formulation ---

1.  Descriptive analysis: what variables accounting for behavioral
    choices and environmental constraints are associated with the final
    grade in Portuguese and Mathematics and how strong are these
    associations.

2.  Predictive analysis: which of the two models, where the first is
    fitted only on structural and environmental constraints and the
    second is fitted on both the constraints and the behavioral choices
    has bigger predictive power.

### Sample Population

Students aged between 15 and 22 that are enrolled in Mathematics and
Portuguese language courses at two distinct schools (Gabriel Pereira and
Mousinho da Silveira).

### Variables

Based on our research question, we have selected the following variables
from both of the datasets for Mathematics and Portuguese (the same
variables are used in both datasets):

**Outcome:**

-   `G3` --- final grade (numeric: from 0 to 20, output target).
    -   *Motivation:* This is the definitive measure of academic success
        for our analysis.

**General:**

-   `sex` --- student's sex (binary: "F" - female or "M" - male).

    -   *Motivation:* Included as a standard demographic control.

-   `age` --- student's age (numeric: from 15 to 22).

    -   *Motivation:* Included as a standard demographic control.

-   `school` --- student's school (binary: "GP" - Gabriel Pereira or
    "MS" - Mousinho da Silveira).

    -   *Motivation:* Included as an environmental control.

**Behavioral Predictors (Choices and Habits) --- Something that a
student can control:**

-   `goout` --- going out with friends (ordinal: from 1 -- very low to 5
    -- very high).

    -   *Motivation:* Represents a student-driven behavioral choice
        regarding time allocation and social priorities.

-   `activities` --- extra-curricular activities (binary: yes or no).

    -   *Motivation:* Reflects proactive personal agency in seeking
        non-academic additions and skill-building.

-   `Dalc*` --- workday alcohol consumption (ordinal: from 1 -- very low
    to 5 -- very high).

    -   *Motivation:* Captures a potentially detrimental personal
        lifestyle choice that directly impacts daily cognitive function
        and academic readiness.

-   `Walc*` --- weekend alcohol consumption (ordinal: from 1 -- very low
    to 5 -- very high).

    -   *Motivation:* Similar to `Dalc`, this isolates weekend social
        choices that might compete with weekend study or recovery time.

-   `studytime` --- weekly study time (ordinal: 1 -- \<2 hours, 2 -- 2
    to 5 hours, 3 -- 5 to 10 hours, or 4 -- \>10 hours).

    -   *Motivation:* The most direct proxy for personal agency, effort,
        and academic discipline.

-   `romantic` --- with a romantic relationship (binary: yes or no).

    -   *Motivation:* Represents a personal commitment that requires an
        allocation of emotional energy and time, potentially competing
        with academic focus.

**Structural/Environmental Predictors (Background and Constraints) ---
something that a student cannot control:**

-   `address` --- student's home address type (binary: "U" -- urban or
    "R" -- rural).

    -   *Motivation:* Captures geographic structural constraints, such
        as access to urban resources, libraries, or potential rural
        isolation.

-   `famsize` --- family size (binary: "LE3" -- less or equal to 3 or
    "GT3" -- greater than 3).

    -   *Motivation:* An environmental factor that may influence the
        home study environment.

-   `Pstatus` --- parent's cohabitation status (binary: "T" -- living
    together or "A" -- apart).

    -   *Motivation:* Reflects foundational family structure and
        potential home-life stability, which is entirely outside the
        student's control.

-   `Medu**` --- mother's education (ordinal: 0 -- none, 1 -- primary
    education (4th grade), 2 –- 5th to 9th grade, 3 –- secondary
    education or 4 –- higher education).

    -   *Motivation:* A proxy for socioeconomic status and inherited
        cultural capital within the home environment.

-   `Fedu**` --- father's education (ordinal: 0 -- none, 1 -- primary
    education (4th grade), 2 –- 5th to 9th grade, 3 –- secondary
    education or 4 – higher education).

    -   *Motivation:* Similar to `Medu`. Complements mother's education
        to fully capture the structural intellectual and socioeconomic
        background provided by the family.

-   `traveltime` --- home to school travel time (ordinal: 1 -- \<15
    min., 2 -- 15 to 30 min., 3 -- 30 min. to 1 hour, or 4 -- \>1 hour).

    -   *Motivation:* A physical, environmental constraint that dictates
        how much "dead time" a student experiences, limiting their
        available hours for study or rest.

-   `internet` --- internet access at home (binary: yes or no)

    -   *Motivation:* Defines access to online-based study materials.

-   `health` --- current health status (ordinal: from 1 -- very bad to 5
    -- very good).

    -   *Motivation:* An inherent biological constraint that affects a
        student's baseline capacity to perform, regardless of their
        willpower or choices.

Note that variables marked with \* and \*\* respectively are
hypothesized to be strongly associated, but are still included in the
initial set with the possibility to be ruled out after Initial Data
Analysis.

### Excluded Variables and Rationale

Remaining variables available in the dataset were intentionally excluded
to maintain a focus on testing our core hypothesis. Past academic
performance metrics (`G1`, `G2`, and past `failures`) were removed;
including these would artificially overwhelm model accuracy.
Additionally, family metrics (like `Mjob`, `Fjob`, and `guardian`) were
excluded because we concluded that the proxy for sociometric status
(Fedu and Medu) were enough and because Mjob and Fjob are just factors
with type of the job (nominal: "teacher", "health" care related, civil
"services" (e.g. administrative or police), "at_home" or "other"), they
do not provide any insight in e.g. family capital or other measurable
metric. Variables (`schoolsup`, `famsup`, `paid`, `nursery`,
`higher`,`freetime` `famrel`) were omitted because we could not strictly
classify them into one of the classes, e.g. attending paid lessons is
based on desire of the student or parents.

### Statistical Approach

To address our research questions, we will use multiple linear
regression. We will fit **four models in total** --- two for the
Mathematics dataset and two for the Portuguese dataset.

#### Model Specifications

For each subject, we define two nested models:

1.  **Structural Model (Base):** This model includes only the inherent
    demographic controls and constraints.
    $$G3 = \beta_0 + \sum_{j \in \text{General}} \beta_j X_j + \sum_{k \in \text{Structural}} \beta_k X_k + \epsilon$$

2.  **Full Model (Agency + Environment):** This model adds behavioral
    predictors to the structural model.
    $$G3 = \beta_0 + \sum_{j \in \text{General}} \beta_j X_j + \sum_{k \in \text{Structural}} \beta_k X_k + \sum_{l \in \text{Behavioral}} \beta_l X_l + \epsilon$$

#### Evaluation Framework

We will evaluate these models through two lenses:

-   Descriptive Analysis & Interpretation: We will analyze and interpet
    the coefficients of both models for each subject to determine which
    specific behaviors or structural factors hold the most weight and
    how they change when within the two models.

-   Model Comparison & Predictive Power: To determine if adding
    "Personal Agency" variables improves our ability to predict academic
    success, we will compare the Structural and Full models using
    diagnostic plots and three metrics:

    -   AIC (Akaike Information Criterion): To assess model fit.
    -   RMSE (Root Mean Square Error), MAE (Mean Absolute Error): To
        provide measures of how close predictions are to actual grades.
    -   CV-RMSE, CV-MAE: Leave-one-out cross-validated equivalents of
        the above, computed via the hat-matrix shortcut exact for OLS.

### Handling Ordinal Factors

Ordinal predictors are encoded in two different ways.

`Medu`, `Fedu`, `studytime` and `traveltime` are kept as ordered factors
with successive difference contrasts. For parental education, the
categories represent qualitatively distinct thresholds (e.g., no
education vs. primary is a fundamentally different jump than secondary
vs. higher), and assuming equal spacing between levels is not justified.
For studytime and traveltime, the categories represent non-uniform time
brackets, and a linear encoding would impose an arbitrary metric on a
coarsely measured constraint.

The remaining ordinal variables --- `goout`, 'Dalc', `Walc`, and
`health` --- are treated as numeric. These are all symmetric 5-point
Likert-style scales. Treating each as a single numeric predictor is also
motivated practically by sample size: 370 (reduced to in IDA 327
observations). Encoding each Likert variable as a factor would consume
an additional 8–10, which we cannot afford with this data.

## Data Management

### Import

```{r}
#| label: data_import

d_mat = read.table("data/student-mat.csv", sep = ";", header = TRUE)
d_por = read.table("data/student-por.csv", sep = ";", header = TRUE)

cat("Number of students of Mathematics:", nrow(d_mat), "\n")
cat("Number of students of Portuguese:", nrow(d_por), "\n")

```

### Merge

We work on both Mathematics and Portuguese --- so we merged them.

```{r}
#| label: data_merge

# NOTE: we used bigger set of variables than in the description of the 
# seminar,because we saw that in the case of using the smaller set, we get 
# non-consistent rows, so two rows in mat and por with the same signature
# gave different responses to e.g. activities
d_combined = merge(d_mat, d_por, 
                   by = c("school", "sex", "age", "address", "famsize",
                          "Pstatus", "Medu", "Fedu", "Mjob", "Fjob",
                          "reason", "guardian", "traveltime", "studytime",
                          "activities", "nursery", "higher", "internet",
                          "romantic", "famrel", "freetime", "goout",
                          "Dalc", "Walc", "health"),
                   suffixes = c(".mat", ".por"))

cat("Number of overlapping students:", nrow(d_combined), "\n")

```

### Variable Cleaning & Recoding

```{r}
#| label: data_transform

# Binary factors
binary_map = list(
  school = list(
    lv = c("GP", "MS"),
    lb = c("0 - Gabriel Pereira", "1 - Mousinho da Silveira")),
  sex = list(
    lv = c("F", "M"),
    lb = c("0 - Female", "1 - Male")),
  address = list(
    lv = c("U", "R"),
    lb = c("0 - Urban", "1 - Rural")),
  famsize = list(
    lv = c("LE3", "GT3"),
    lb = c("0 - <=3 members", "1 - >3 members")),
  Pstatus = list(
    lv = c("A", "T"),
    lb = c("0 - Apart", "1 - Living together")),
  activities = list(
    lv = c("no", "yes"),
    lb = c("0 - No", "1 - Yes")),
  romantic = list(
    lv = c("no", "yes"),
    lb = c("0 - No", "1 - Yes")),
  internet = list(
    lv = c("no", "yes"),
    lb = c("0 - No", "1 - Yes")))

for (v in names(binary_map)) {
  d_combined[[v]] = factor(
    d_combined[[v]],
    levels = binary_map[[v]]$lv,
    labels = binary_map[[v]]$lb)
}

# Ordered factors 
edu_levels = c(
  "0 - None",
  "1 - Primary (4th grade)",
  "2 - 5th-9th grade",
  "3 - Secondary",
  "4 - Higher education"
)

d_combined$Medu = factor(
  d_combined$Medu,
  levels = 0:4,
  labels = edu_levels,
  ordered = TRUE)

d_combined$Fedu = factor(
  d_combined$Fedu,
  levels = 0:4,
  labels = edu_levels,
  ordered = TRUE)

d_combined$traveltime = factor(
  d_combined$traveltime,
  levels = 1:4,
  labels = c(
    "1 - <15 min",
    "2 - 15-30 min",
    "3 - 30 min-1 h",
    "4 - >1 h"),
  ordered = TRUE)

d_combined$studytime = factor(
  d_combined$studytime,
  levels = 1:4,
  labels = c(
    "1 - <2 h",
    "2 - 2-5 h",
    "3 - 5-10 h",
    "4 - >10 h"),
  ordered = TRUE)

# Define variable groups and subset to analytic datasets
general_vars  = c("school", "sex", "age")
behavioural_vars = c("goout", "activities", "Dalc", 
                     "Walc", "studytime", "romantic")
structural_vars = c("address", "famsize", "Pstatus", 
                    "Medu", "Fedu", "traveltime", "internet", "health")

# List of all ordinal variables that need incremental coding
ordinal_vars = c("Medu", "Fedu", "traveltime", "studytime")

# Apply Successive Difference Contrasts
for (v in ordinal_vars) {
  num_levels = length(levels(d_combined[[v]]))
  contrasts(d_combined[[v]]) = MASS::contr.sdif(num_levels)
}

prepare_subject_data = function(df, subject_suffix) {
  shared_cols = c(general_vars, behavioural_vars, structural_vars)
  df_sub = df[, c(shared_cols, paste0("G3", subject_suffix))]
  colnames(df_sub) = gsub(subject_suffix, "", colnames(df_sub), fixed = TRUE)
  return(df_sub)
}

d_mat_analysis = prepare_subject_data(d_combined, ".mat")
d_por_analysis = prepare_subject_data(d_combined, ".por")

```

## Initial Data Analysis

### IDA Screening: Missing Values

```{r}
#| label: ida_missing_values

shared_vars_ida = setdiff(names(d_mat_analysis), "G3")
d_unified_ida = d_mat_analysis[, shared_vars_ida]
d_unified_ida$G3_Math = d_mat_analysis$G3
d_unified_ida$G3_Por = d_por_analysis$G3

d_unified_ida |>
  miss_var_summary() |>
  gt() |>
  tab_header(
    title = "Missing Value Summary",
    subtitle = "Predictor variables are shared across subjects; 
    G3 shown separately per subject")

```

There are no NAs in this dataset.

#### Zero-values (e.g. None in Parent's education)

These are valid encodings, but they are special compared to other values
--- so we needed to look closer at them.

```{r}
#| label: ida_zero_check

zero_summary = data.frame(
  Variable = c("Medu = None", "Fedu = None",
                "G3 = 0 (Math)", "G3 = 0 (Portuguese)"),
  N = c(sum(d_mat_analysis$Medu == "0 - None"),
        sum(d_mat_analysis$Fedu == "0 - None"),
        sum(d_mat_analysis$G3 == 0),
        sum(d_por_analysis$G3 == 0)))

zero_summary |>
  gt() |>
  tab_header(title = "Zero / Baseline Value Counts", 
             subtitle = "Medu & Fedu: parents with no education; 
             G3: students scoring 0") |>
  cols_label(Variable = "Variable", N = "n")

```

The first observation we could make is that for Medu = None and Fedu =
None looking at the records, we could not understand whether these
really correspond to not having education or they encode something else.
Given that the presence of both is very small in the sample, which would
even if the encodings are truthful not produce meaningful results (we
used incremental contrast factors for these variables) in the model, we
decided to exclude these records.

```{r}
#| label: data_exclusions

d_combined = d_combined |>
  filter(Medu != "0 - None", Fedu != "0 - None")

d_combined$Medu = droplevels(d_combined$Medu)
d_combined$Fedu = droplevels(d_combined$Fedu)

contrasts(d_combined$Medu) = MASS::contr.sdif(length(levels(d_combined$Medu)))
contrasts(d_combined$Fedu) = MASS::contr.sdif(length(levels(d_combined$Fedu)))

cat("Observations after exclusion:", nrow(d_combined), "\n")

```

The second observation is for G3 being 0, there is no specification in
student.txt so we were not sure what is the meaning. We supposed that
this corresponds to dropouts from the course, which if true would
largely inflate our models; because we wanted to focus on the difference
in final grades not whether the student finishes the course. To check
the assumption that this corresponds to dropouts, we summarized these
cases below.

```{r}
#| label: ida_zero_g3_investigation

# Pull G1, G2 from d_combined for analytic dataset students 
g_mat = d_combined |> select(G1 = G1.mat, G2 = G2.mat, G3 = G3.mat)
g_por = d_combined |> select(G1 = G1.por, G2 = G2.por, G3 = G3.por)

g_mat |> filter(G3 == 0) |> as_tibble() |> print(n = Inf)
g_por |> filter(G3 == 0) |> as_tibble() |> print(n = Inf)

# Predictor means: G3=0 vs G3>0
make_zero_summary = function(df, subject) {
  df |>
    mutate(zero_g3 = ifelse(G3 == 0, "G3 = 0", "G3 > 0")) |>
    group_by(zero_g3) |>
    summarise(n = n(), mean_age = mean(age),
              mean_studytime = mean(as.numeric(studytime)),
              mean_goout = mean(as.numeric(goout)),
              mean_Dalc = mean(as.numeric(Dalc)),
              mean_Walc = mean(as.numeric(Walc)),
              mean_Medu = mean(as.numeric(Medu)),
              mean_Fedu = mean(as.numeric(Fedu))) |> gt() |>
    tab_header(title = 
                 paste("Predictor Means: G3 = 0 vs G3 > 0 —", subject)) |>
    fmt_number(columns = where(is.double), decimals = 2)
}

make_zero_summary(d_mat_analysis, "Mathematics")
make_zero_summary(d_por_analysis, "Portuguese")

```

First, in the Mathematics dataset, we observed that a subset of students
with a final grade of zero (G3 = 0) also lack recorded grades for the
second period (G2). Additionally, among their observable characteristics
there is nothing indicating that these are some special group of
students --- there are differences especially among Portuguese students;
but they are just other indicators that these students are dropouts
(e.g. higher alcohol consumption --- which is the characteristic that,
we think, is generally agreed to be associated with not finishing some
duty -\> dropping out). So, this pattern suggests that these cases most
likely correspond to students who dropped out. As we discussed before,
keeping them would distort the analysis, so, we excluded these
observations.

```{r}
#| label: symmetric_exclusion

d_combined = d_combined |> 
  filter(G3.mat > 0, G3.por > 0)

n_after_sym = nrow(d_combined)

cat("Observations after (G3 > 0 in both):", n_after_sym, "\n")

```

Since we excluded the students from the combined datasets, we needed to
regenerate the analytic datasets with excluded observations.

```{r}
#| label: regenerate_analytic_data

d_mat_analysis = prepare_subject_data(d_combined, ".mat")
d_por_analysis = prepare_subject_data(d_combined, ".por")

```

### IDA Screening: Univariate Descriptions

In this subsection we summarize univariate characteristics of the
variables we selected.

#### Continuous Variables

```{r}
#| label: ida_continuous_plots
#| fig-width: 12
#| fig-height: 4

p_age = ggplot(d_mat_analysis, aes(x = age)) +
  geom_histogram(binwidth = 1,
                 fill = viridis(1, option = VIR_PAL, begin = 0.2),
                 colour = "white") +
  geom_vline(xintercept = mean(d_mat_analysis$age), linetype = "dashed") +
  labs(title = "Age", x = "Age (years)", y = "Count")

p_g3_mat = ggplot(d_mat_analysis, aes(x = G3)) +
  geom_histogram(binwidth = 1,
                 fill = viridis(1, option = VIR_PAL, begin = 0.5),
                 colour = "white") +
  geom_vline(xintercept = mean(d_mat_analysis$G3), linetype = "dashed") +
  labs(title = "Final Grade - Mathematics", x = "G3 (1-20)", y = "Count")

p_g3_por = ggplot(d_por_analysis, aes(x = G3)) +
  geom_histogram(binwidth = 1,
                 fill = viridis(1, option = VIR_PAL, begin = 0.8),
                 colour = "white") +
  geom_vline(xintercept = mean(d_por_analysis$G3), linetype = "dashed") +
  labs(title = "Final Grade - Portuguese", x = "G3 (1-20)", y = "Count")

(p_age | p_g3_mat | p_g3_por) +
  plot_annotation(
    title    = "Univariate Distributions - Continuous Variables",
    theme    = theme(plot.title = element_text(size = 14,
                                               face = "bold", hjust = 0.5)))

```

```{r}
#| label: ida_continuous_hmisc

cont_df = data.frame(Age = d_mat_analysis$age, G3_Math = d_mat_analysis$G3,
                     G3_Portuguese = d_por_analysis$G3)

s_cont = Hmisc::describe(cont_df, digits = 1)
attr(s_cont, "descript") = "Continuous variables - overall characteristics"
Hmisc::html(s_cont)

```

Below is the summary of findings for Contionous variables:

For **Age**, the distribution is concentrated in the 15–18 range, which
together account for 97.5% of the sample. The mean and median both sit
at 16–17 years, as expected for secondary school students. The tail
extends to age 22, with two isolated observations at ages 20 and 22 (one
student each). The single 22-year-old is a potential anomaly, but in the
documentation the range is (18-22), so we kept it.

For **G3 - Mathematics**, the distribution is roughly symmetric around
the median and mean of 12 and grades ranging from 4 to 20. The most
common grade is 10 (16.2% of students), and the interquartile range
spans 9-14, suggesting moderate spread.

For **G3 - Portuguese**, the distribution is more concentrated than
Mathematics, with a mean and median both at 13 and a narrower IQR of
11–15. The distribution is approximately symmetric but with moderate
left-skewedness. There is one isolated observation at grade 1, which is
notably lower than the next minimum of 5 --- this single case grabs
attention, but given that we already excluded G3 = 0 as dropouts, we
kept it.

#### Categorical Variables

Note that the value of variables for Mathematics and Portuguese are the
same. So below we use data for Mathematics, but it is completely similar
for Portuguese.

```{r}
#| label: ida_ordinal_plots
#| fig-width: 14
#| fig-height: 9

pal_ord = viridis(8, option = VIR_PAL, begin = 0.1, end = 0.9)

# Labeled copy for plotting only — does not affect analytic datasets
d_mat_plot = d_mat_analysis |>
  mutate(goout = factor(goout, levels = 1:5,
                        labels = c("1 - very low", "2", "3", 
                                   "4", "5 - very high")),
         Dalc = factor(Dalc, levels = 1:5,
                       labels = c("1 - very low", "2", "3",
                                  "4", "5 - very high")),
         Walc = factor(Walc, levels = 1:5,
                       labels = c("1 - very low", "2", "3", 
                                  "4", "5 - very high")),
         health = factor(health, levels = 1:5,
                       labels = c("1 - very bad", "2", "3", 
                                  "4", "5 - very good")))

# Reusable function to make bar plots
make_bar = function(var, title, fill_col) {
  ggplot(d_mat_plot, aes(x = .data[[var]])) +
    geom_bar(fill = fill_col, colour = "white", width = 0.7) +
    geom_text(stat = "count", aes(label = after_stat(count)),
              vjust = -0.4, size = 3.2, colour = "grey20") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1),
          panel.grid.major.x = element_blank()) +
    labs(title = title, x = NULL, y = "Count")
}

p_medu = make_bar("Medu", "Mother's Education", pal_ord[1])
p_fedu = make_bar("Fedu", "Father's Education", pal_ord[2])
p_tt = make_bar("traveltime", "Travel Time", pal_ord[3])
p_st = make_bar("studytime", "Weekly Study Time", pal_ord[4])
p_go = make_bar("goout", "Going Out with Friends", pal_ord[5])
p_dalc = make_bar("Dalc", "Workday Alcohol Consumption", pal_ord[6])
p_walc = make_bar("Walc", "Weekend Alcohol Consumption", pal_ord[7])
p_hlth = make_bar("health", "Health Status", pal_ord[8])

(p_medu | p_fedu | p_tt | p_st) /
(p_go | p_dalc | p_walc | p_hlth) +
  plot_annotation(
    title = "Univariate Distributions - Ordinal Variables",
    theme = theme(plot.title = element_text(size = 14,
                                            face = "bold", hjust = 0.5)))

```

Below is the summary of findings for Categorical variables:

**Medu and Fedu**: In the case of mother's education, the distribution
is somewhat left-skewed, whereas the distribution for father's education
is closer to uniform. This observation was kept until multivariate
analysis, where one variable may be prioritized.

**traveltime**: The distribution is right-skewed. The top two levels (16
and 7 observations) were merged into a single "\>=30 min" category
before model fitting, as small cell sizes in these strata would produce
unstable coefficient estimates.

**studytime**: The distribution is reasonably spread, with the majority
studying 2–5 hours per week. The upper level (\>10 hours) is the
smallest group with only 23 students - we thought of merging the two top
categories, but in the end decided not to because we supposed that this
is most likely one of the strongest predictors, and the strata size is
small but not extremely.

**goout**: The distribution is close to symmetric and centered on level
3. Level 1 ("very low") has only 18 students, which is notably sparse
relative to the remaining levels, but since we code it as numeric, this
should not play a significant role.

**Dalc and Walc**: These distributions differ notably; Dalc is heavily
concentrated at "very low" (approximately 70.0%), with very sparse
counts at levels 4 and 5. Walc is more spread across all levels, with
counts for levels 4 and 5 still somewhat underrepresented but sufficient
to retain as separate levels. Given their expected collinearity, the
decision on retaining or excluding either variable was left until
multivariate analysis.

**health**: The distribution is left-skewed. All five levels have
reasonable counts.

```{r}
#| label: ida_binary_plots
#| fig-width: 14
#| fig-height: 4.5

pal_bin = viridis(8, option = VIR_PAL, begin = 0.1, end = 0.9)

p_sch = make_bar("school", "School", pal_bin[1])
p_sex = make_bar("sex", "Sex", pal_bin[2])
p_adr = make_bar("address", "Home Address", pal_bin[3])
p_fsz = make_bar("famsize", "Family Size", pal_bin[4])
p_pst = make_bar("Pstatus", "Parental Status", pal_bin[5])
p_act = make_bar("activities", "Extra-curriculars", pal_bin[6])
p_rom = make_bar("romantic", "Romantic Relationship", pal_bin[7])
p_net = make_bar("internet", "Internet at Home", pal_bin[8])

(p_sch | p_sex | p_adr | p_pst) /
(p_fsz | p_act | p_rom | p_net) +
  plot_annotation(
    title = "Univariate Distributions - Binary Variables",
    theme = theme(plot.title = element_text(size = 14,
                                            face = "bold", hjust = 0.5)))

```

Below is the summary of findings for Binary variables:

For **school**, the sample is heavily dominated by Gabriel Pereira, with
only 32 students from Mousinho da Silveira. This imbalance is worth
keeping in mind as the school coefficient will be estimated from a very
small contrast group.

For **sex**, the split is nearly equal.

For **address**, the majority of students live in urban area, with rural
students representing about a fifth of the sample --- both groups are
sufficiently represented.

For **Pstatus and famsize**, both variables show imbalance in the same
direction: most students come from larger families and from households
where parents live together. The "Apart" category for Pstatus is
particularly sparse, which may produce an unstable estimate --- we kept
this in mind.

For **activities and romantic**, activities is nearly balanced. Romantic
relationships are less common but still well-represented, with a little
bit less than third of students reporting being in a relationship.

For **internet**, the large majority of students have internet access at
home, leaving only 49 without. This imbalance is not extreme enough to
require action, but we kept it in mind.

### IDA Screening: Multivariate Descriptions

In this subsection we summarize multivariate characteristics of the
variables we selected.

We performed stratification on two general variables, namely, by **sex**
and by **school**. Stratification by all remaining variables was omitted
to avoid inflating the report with uninformative visuals --- the
correlation structure between predictors will be addressed more
systematically in the subsequent correlation and redundancy analysis.

#### Stratification by Sex

```{r}
#| label: ida_strat_sex_plots
#| fig-width: 12
#| fig-height: 8

p1 = ggplot(d_mat_analysis, aes(x = sex, y = age)) +
  geom_boxplot() + theme_bw() +
  labs(x = "Sex", y = "Age (years)")

p2 = ggplot(d_mat_analysis, aes(x = studytime, fill = sex)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = VIR_PAL, begin = 0.2, 
                       end = 0.7, name = "Sex") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = "Study Time", y = "Proportion")

p3 = ggplot(d_mat_plot, aes(x = goout, fill = sex)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = VIR_PAL, begin = 0.2,
                       end = 0.7, name = "Sex") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = "Going Out", y = "Proportion")

p4 = ggplot(d_mat_plot, aes(x = Walc, fill = sex)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = VIR_PAL, begin = 0.2, 
                       end = 0.7, name = "Sex") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = "Weekend Alcohol", y = "Proportion")

(p1) / (p2 | p3 | p4) +
  plot_annotation(
    title = "Predictors Stratified by Sex",
    theme = theme(plot.title = element_text(size = 14, 
                                            face = "bold", hjust = 0.5)))

```

Notable observations are: that female student spent more time studying,
and consume smaller amount of alcohol on weekends compared to male
students in this sample; especially when it comes to higher amounts of
consumption. In terms of age female students are a little bit older and
their age is less spread out compared to male students.

#### Stratification by School

```{r}
#| label: ida_strat_school_plots
#| fig-width: 12
#| fig-height: 8

rot_x = theme(axis.text.x = element_text(angle = 30, hjust = 1))

p1 = ggplot(d_mat_analysis, aes(x = school, y = age)) +
  geom_boxplot() + theme_bw() +
  labs(x = "School", y = "Age (years)")

p2 = ggplot(d_mat_analysis, aes(x = school, fill = Fedu)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = VIR_PAL, name = "Father's Edu") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  rot_x +
  labs(x = "School", y = "Proportion")

p3 = ggplot(d_mat_analysis, aes(x = school, fill = Medu)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = VIR_PAL, name = "Mother's Edu") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  rot_x +
  labs(x = "School", y = "Proportion")

p4 = ggplot(d_mat_analysis, aes(x = school, fill = address)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = VIR_PAL, begin = 0.2,
                       end = 0.7, name = "Address") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  rot_x +
  labs(x = "School", y = "Proportion")

(p1) / (p2 | p3 | p4) +
  plot_annotation(
    title = "Predictors Stratified by School",
    theme = theme(plot.title = element_text(size = 14,
                                            face = "bold", hjust = 0.5)))

```

Notable observations are: that students from Mousinho da Silveira are
generally older, and their parents --- particularly their mothers ---
have notably lower formal education levels compared to the families at
Gabriel Pereira. Furthermore, there is a geographic divide, with half of
Mousinho da Silveira's students living in rural areas, whereas Gabriel
Pereira serves a predominantly urban population. Overall, Mousinho da
Silveira reflects a demographic facing more pronounced environmental and
socioeconomic constraints.

#### Correlation Between All Predictors

We now summarize all of the correlations.

```{r}
#| label: ida_heatmap_all
#| fig-width: 9
#| fig-height: 8

# Make a numeric-only dataframe where factors are converted to ranks
cor_all_df = d_mat_analysis |>
  mutate(School = as.numeric(school),
         Sex = as.numeric(sex),
         Age = age,
         Address = as.numeric(address),
         Famsize = as.numeric(famsize),
         Pstatus = as.numeric(Pstatus),
         Medu = as.numeric(Medu),
         Fedu = as.numeric(Fedu),
         Traveltime = as.numeric(traveltime),
         Internet = as.numeric(internet),
         Studytime = as.numeric(studytime),
         Activities = as.numeric(activities),
         Romantic = as.numeric(romantic),
         Goout = as.numeric(goout),
         Dalc = as.numeric(Dalc),
         Walc = as.numeric(Walc),
         Health = as.numeric(health)) |>
  select(School, Sex, Age, Address, Famsize, Pstatus, Medu, Fedu, 
         Traveltime, Internet, Studytime, Activities, Romantic, 
         Goout, Dalc, Walc, Health)

# Use Spearman correlation
cor_matrix = cor(cor_all_df, use = "pairwise.complete.obs", 
                 method = "spearman")

Breaks = seq(-1, 1, by=.01)
my_palette = c(colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, 
                                      name ="RdYlBu")))
                (length(Breaks)-2), "grey80", "grey80")

pheatmap(cor_matrix, cluster_cols= TRUE, cluster_rows = TRUE,
         display_numbers = TRUE, fontsize_number = 7,
         main= "Spearman Correlation Matrix (All Variables)", 
         breaks = Breaks, color = my_palette)

```

We observe some expected clusters. The strongest positive correlations
are between Mother's Education (`Medu`) and Father's Education (`Fedu`),
as well as between Workday Alcohol Consumption (`Dalc`) and Weekend
Alcohol Consumption (`Walc`) - what we hypothesized in the beginning.
There is also a notable positive association between `Walc` and `Goout`
(similarly for `Dalc` and `Goout` but weaker) - which is expectable
given that some of going out activities include alcohol consumption.
Additionaly, there is a negative association between `studytime` and
`sex` - the observation we made when performing stratification by sex.
Additionally, note that Spearman correlation is used as a common ground
for our mix of binary, ordinal, and continuous variables, rather than as
the theoretically optimal measure for every pair.

To make a final decision of keeping or not keeping all of the variables,
we wanted to perform Redundancy Analysis.

#### Redundancy Analysis

Note that we set nk = 0 to suppress spline transformations, as most
predictors have only 4 ordinal levels --- too few to support non-linear
fits. This reduces the redundancy analysis to a linear R\^2 check rather
than a full non-linear redundancy assessment.

```{r}
#| label: ida_redundancy

Hmisc::redun(~ age + school + sex + address + famsize + Pstatus + 
               Medu + Fedu + traveltime + internet + health +
               goout + activities + Dalc + Walc + studytime + romantic, 
             data = d_combined, nk = 0)

```

The redundancy analysis showed that no variable exceeds the R\^2 = 0.9
redundancy threshold, meaning no predictor can be linearly reconstructed
from the remaining variables with sufficient precision to warrant
automatic exclusion. The highest R\^2 values belong to `Walc` (0.554)
and `Dalc` (0.484), reflecting their shared alcohol-consumption
construct --- consistent with the correlation observed in the heatmap.
`Medu` and `Fedu` follow at 0.455 and 0.432 respectively, again expected
given their correlation, but well below the redundancy cutoff. All
remaining variables have R\^2 below 0.33, indicating they each
contribute largely independent information. On these facts alone, no
variable would be excluded --- however, the decisions were revisited in
the IDA conclusions.

### IDA Conclusions

Based on the screening, three sets of decisions are carried into model
fitting.

**Variable exclusion.** `Dalc` is dropped from the analytic models. This
decision is not motivated by multicollinearity (which as we have seen is
present, but not extreme), rather, the justification is distributional:
most of students score at the lowest level, and after collapsing the
sparse upper categories the variable would effectively reduce to a
near-binary indicator. `Walc`, which is better spread across all five
levels and captures the same underlying alcohol consumption construct,
is thus kept as the sole alcohol consumption predictor.

**Level recoding**. For `traveltime`, the top two levels ("30 min –- 1
h" and "\>1 h") are merged into a single "\>=30 min" category, as the
original cell sizes (16 and 7 observations respectively) were too small
to produce meaningful coefficient estimates.

**Underrepresented categories.** Three binary variables show notable
imbalance: `school` (only 32 students from Mousinho da Silveira),
`Pstatus` (36 observations in "Apart" category), and `internet` (only 49
students without home access). These variables are kept in the models,
as their imbalance is not severe enough to motivate exclusion and each
carries plausible theoretical relevance. However, their coefficient
estimates should be interpreted with caution --- the contrast groups are
small, confidence intervals will be wide, and any effects may be
unstable.

```{r}
#| label: ida_recodings

# Drop Dalc from behavioral variable list
behavioural_vars = c("goout", "activities", "Walc", "studytime", "romantic")

# Collapse traveltime: merge levels 3 and 4 into "3 - >=30 min"
d_combined$traveltime = fct_collapse(
  d_combined$traveltime,
  "3 - >=30 min" = c("3 - 30 min-1 h", "4 - >1 h"))
d_combined$traveltime = droplevels(d_combined$traveltime)

# Reapply successive difference contrasts after level change
contrasts(d_combined$traveltime) = MASS::contr.sdif(
  length(levels(d_combined$traveltime)))

# Regenerate analytic datasets with updated variable list and recodings
d_mat_analysis = prepare_subject_data(d_combined, ".mat")
d_por_analysis = prepare_subject_data(d_combined, ".por")

```

### Final Summary Before Analysis

```{r}
#| label: ida_final_summary

var_summary = tribble(
  ~Group, ~Variable, ~Type, ~Levels,
  "General",     "school",     "Binary",  paste(levels(d_combined$school), collapse = " / "),
  "General",     "sex",        "Binary",  paste(levels(d_combined$sex), collapse = " / "),
  "General",     "age",        "Numeric", "15-22",
  "Behavioural", "goout",      "Numeric", "1 (very low) – 5 (very high)",
  "Behavioural", "activities", "Binary",  paste(levels(d_combined$activities), collapse = " / "),
  "Behavioural", "Walc",       "Numeric", "1 (very low) – 5 (very high)",
  "Behavioural", "studytime",  "Ordinal", paste(levels(d_combined$studytime), collapse = " / "),
  "Behavioural", "romantic",   "Binary",  paste(levels(d_combined$romantic), collapse = " / "),
  "Structural",  "address",    "Binary",  paste(levels(d_combined$address), collapse = " / "),
  "Structural",  "famsize",    "Binary",  paste(levels(d_combined$famsize), collapse = " / "),
  "Structural",  "Pstatus",    "Binary",  paste(levels(d_combined$Pstatus), collapse = " / "),
  "Structural",  "Medu",       "Ordinal", paste(levels(d_combined$Medu), collapse = " / "),
  "Structural",  "Fedu",       "Ordinal", paste(levels(d_combined$Fedu), collapse = " / "),
  "Structural",  "traveltime", "Ordinal", paste(levels(d_combined$traveltime), collapse = " / "),
  "Structural",  "internet",   "Binary",  paste(levels(d_combined$internet), collapse = " / "),
  "Structural",  "health",     "Numeric", "1 (very bad) – 5 (very good)",
  "Outcome",     "G3",         "Numeric", "1-20"
)

var_summary |>
  gt(groupname_col = "Group") |>
  tab_header(
    title = "Final Analytic Variable Set",
    subtitle = paste0("n = ", nrow(d_combined), " observations")) |>
  cols_label(Variable = "Variable", Type = "Type", Levels = "Levels / Range") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_row_groups()) |>
  tab_style(style = cell_text(style = "italic"),
            locations = cells_body(columns = Variable))

```

## Model Analysis

### Fit the Models

```{r}
formula_structural = G3 ~ school + sex + age + 
  address + famsize + Pstatus + Medu + Fedu + 
  traveltime + internet + health

formula_full = update(formula_structural,
  . ~ . + goout + activities + Walc + studytime + romantic)

mod_mat_struct = lm(formula_structural, data = d_mat_analysis)
mod_mat_full = lm(formula_full, data = d_mat_analysis)
mod_por_struct = lm(formula_structural, data = d_por_analysis)
mod_por_full = lm(formula_full, data = d_por_analysis)

```

```{r}
#| label: coef_labels

term_labels = c(
  `(Intercept)`  = "Intercept",
  `school1 - Mousinho da Silveira` = "School: Mousinho da Silveira (ref: Gabriel Pereira)",
  `sex1 - Male` = "Sex: Male (ref: Female)",
  `age` = "Age",
  `address1 - Rural` = "Address: Rural (ref: Urban)",
  `famsize1 - >3 members` = "Family Size: >3 (ref: <=3)",
  `Pstatus1 - Living together` = "Parental Status: Together (ref: Apart)",
  `Medu2-1` = "Mother's Edu: 5th-9th vs Primary",
  `Medu3-2` = "Mother's Edu: Secondary vs 5th-9th",
  `Medu4-3` = "Mother's Edu: Higher vs Secondary",
  `Fedu2-1` = "Father's Edu: 5th-9th vs Primary",
  `Fedu3-2` = "Father's Edu: Secondary vs 5th-9th",
  `Fedu4-3` = "Father's Edu: Higher vs Secondary",
  `traveltime2-1` = "Travel: 15-30 min vs <15 min",
  `traveltime3-2` = "Travel: >=30 min vs 15-30 min",
  `internet1 - Yes`= "Internet: Yes (ref: No)",
  `health` = "Health (numeric, 1–5)",
  `goout` = "Going Out (numeric, 1–5)",
  `activities1 - Yes` = "Activities: Yes (ref: No)",
  `Walc` = "Weekend Alcohol (numeric, 1–5)",
  `studytime2-1` = "Study Time: 2-5 h vs <2 h",
  `studytime3-2` = "Study Time: 5-10 h vs 2-5 h",
  `studytime4-3` = "Study Time: >10 h vs 5-10 h",
  `romantic1 - Yes` = "Romantic: Yes (ref: No)"
)

make_coef_table = function(model, title_str) {
  tidy(model, conf.int = TRUE) |>
    mutate(term = coalesce(term_labels[term], term)) |>
    select(Term = term, Estimate = estimate,
           `95% CI lower` = conf.low,
           `95% CI upper` = conf.high,
           `p-value` = p.value) |>
    gt() |>
    tab_header(title = title_str,
               subtitle = "Estimates with 95% Confidence Intervals") |>
    fmt_number(columns = c(Estimate, `95% CI lower`, `95% CI upper`),
               decimals = 3) |>
    fmt_scientific(columns = `p-value`, decimals = 2) |>
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_body(rows = `p-value` < 0.05))
}

```

### Mathematics

In this subsection we will focus on interpretation of the effects for
Mathematics.

#### Coefficients of Structural Model

```{r}
#| label: coef_mat_struct

make_coef_table(mod_mat_struct, "Mathematics - Structural Model")

```

#### Coefficients of Full Model

```{r}
#| label: coef_mat_full

make_coef_table(mod_mat_full, "Mathematics - Full Model")

```

#### Interpretation of Coefficients

##### Structural Model

Only two structural predictors reach significance. **Sex** is the
clearest signal: male students score 0.94 points higher on average (95%
CI: 0.23 to 1.65). **Health** is the other, and its direction is
counterintuitive - each unit increase on the 1–5 scale (bigger value the
better is reported health) is associated with a 0.28 point *decrease* in
grade (95% CI: -0.53 to -0.03). All remaining structural predictors ---
parental education, travel time, internet, address, family size, school
—-- are non-significant, suggesting that environmental background
explains very little of the variation in mathematics performance on its
own.

##### Full Model

**Sex** remains significant and grows slightly: male students score 1.27
points higher (95% CI: 0.54 to 2.01). The fact that the gap widens after
accounting for study time is counterintuitive. Female students in the
IDA were observed to study more, so one might expect the gap to shrink.
Instead, controlling for study time makes the gap larger, suggesting
that these behavioral differences were partially offsetting the
underlying grade gap rather than explaining it. **Health** retains the
same negative estimate (−0.28, 95% CI: -0.52 to -0.03); we suspected
confounding with social lifestyle, but the consistency across both
models suggests something else (the explanation was found in
non-linearity analysis).

Among behavioral predictors, **study time** shows the clearest effect,
though the pattern is non-monotone across contrasts. The jump from 2–5
hours to 5–10 hours per week is the strongest single effect in the model
(+1.52 points, 95% CI: 0.53 to 2.52), while the contrasts involving the
lowest and highest study groups are small and non-significant. This
points to a threshold effect: moving from moderate to substantial weekly
study time matters, but studying beyond that does not yield further
gains.

**Going out** is the one other behavioral predictor that is significant:
each additional unit on the 1–5 scale is associated with a 0.38 point
decrease in grade (95% CI: −0.72 to −0.03). **Weekend alcohol** is
borderline (−0.30, 95% CI: −0.62 to 0.01) and we would not claim it as a
finding, though its direction is consistent with the broader pattern of
social-behavioral factors pulling grades down. **Romantic relationship**
and **activities** are both non-significant with near-zero estimates.

Note that the intercept is not interpreted. With successive difference
contrasts applied to the ordered factors (parental education, travel
time, study time), the intercept represents the mean across factor
levels rather than a value at any natural reference category. It does
not correspond to an interpretable student profile.

### Portuguese

In this subsection we focus on Portuguese, but also connecting it to
results of Mathematics.

#### Coefficients of Structural Model

```{r}
#| label: coef_por_struct

make_coef_table(mod_por_struct, "Portuguese - Structural Model")

```

#### Coefficients of Full Model

```{r}
#| label: coef_por_full

make_coef_table(mod_por_full, "Portuguese - Full Model")

```

#### Interpretation of Coefficients

##### Structural Model

As in Mathematics, **health** is significant and negative (−0.23, 95%
CI: -0.41 to -0.22). **Sex** is again the clearest structural signal,
but the direction reverses: female students score 0.97 points *higher*
in Portuguese (95% CI: 0.42 to 1.52), compared to male students scoring
higher in Mathematics.

Parental education is non-significant, with one exception: the contrast
between higher education and secondary for mother's education reaches
significance (+0.88, 95% CI: 0.13 to 1.64). Unlike for Mathematics this
suggests that the benefit of parental education for Portuguese grades is
significant but just compared to parents having higher degree and having
secondary school degree. All remaining structural predictors — school,
family size, parental status, travel time, internet, address — are
non-significant.

##### Full Model

**Sex** remains significant after adding behavioral predictors, though
the estimate goes down to 0.61 (95% CI: 0.05 to 1.17), suggesting
behavioral differences between sexes explain part of the gap in the case
of Portuguese. **Age** becomes significant in the full model (+0.35 per
year, 95% CI: 0.1 to 0.6), which was not the case in the structural
model - this is likely a suppression effect, where behavioral variables
that correlate with age were masking its association with the grade.
**Address** also crosses the significance threshold in the full model:
rural students score 0.94 points lower on average (95% CI: −1.65 to
−0.24), a pattern absent in Mathematics. **Health** keeps its negative
estimate (−0.22, 95% CI: −0.41 to -0.03), again consistent with
Mathematics. The mother's education contrast (Higher vs Secondary)
remains significant at +0.83 (95% CI: 0.11 1.55), and father's education
at the same contrast now also reaches significance (+0.76, p = 0.047)
--- a stronger parental education signal than in Mathematics overall.

Among behavioral predictors, **going out** is significant and negative
(−0.43 per unit, 95% CI: −0.69 to −0.16), consistent with Mathematics in
both direction and magnitude. **Study time** again shows a non-monotone
pattern, but smaller compared to Mathematics: the 5–10 h vs 2–5 h
contrast is significant (+0.84, 95% CI: 0.01 to 1.6), while the lowest
and highest contrasts are not —-- the same threshold structure as in
Mathematics, but with a smaller effect at the critical step. **Weekend
alcohol**, **romantic relationship**, and **activities** are all
non-significant, mirroring the Mathematics results.

The overall picture in comparison with Mathematics: structural
background is weak, the sex gap persists but flips direction, and the
behavioral signal is again concentrated in study time intensity and
going out. The notable differences are that Portuguese grades show a
stronger parental education effect at the upper end, a rural address
penalty absent in Mathematics, and an age effect that only emerges once
behavioral variables are controlled for.

### Model Comparison

In this subsection we compare the predictive power of the models.

```{r}
#| label: model_comparison

rmse = function(model) sqrt(mean(resid(model)^2))
mae = function(model) mean(abs(resid(model)))

# Exact LOOCV via leverage shortcut: e_(-i) = e_i / (1 - h_ii)
loocv_rmse = function(model) {
  e = resid(model); h = hatvalues(model)
  sqrt(mean((e / (1 - h))^2))
}
loocv_mae = function(model) {
  e = resid(model); h = hatvalues(model)
  mean(abs(e / (1 - h)))
}

comparison = tibble(
  Subject = c("Mathematics", "Mathematics", "Portuguese", "Portuguese"),
  Model = c("Structural", "Full", "Structural", "Full"),
  
  AIC = c(AIC(mod_mat_struct), AIC(mod_mat_full),
          AIC(mod_por_struct), AIC(mod_por_full)),
  
  RMSE = c(rmse(mod_mat_struct), rmse(mod_mat_full),
           rmse(mod_por_struct), rmse(mod_por_full)),
  
  MAE = c(mae(mod_mat_struct), mae(mod_mat_full),
          mae(mod_por_struct), mae(mod_por_full)),
  CV_RMSE = c(loocv_rmse(mod_mat_struct), loocv_rmse(mod_mat_full),
              loocv_rmse(mod_por_struct), loocv_rmse(mod_por_full)),
  
  CV_MAE= c(loocv_mae(mod_mat_struct), loocv_mae(mod_mat_full),
            loocv_mae(mod_por_struct), loocv_mae(mod_por_full)),
  
  R2 = c(summary(mod_mat_struct)$r.squared,
         summary(mod_mat_full)$r.squared,
         summary(mod_por_struct)$r.squared,
         summary(mod_por_full)$r.squared),
  
  Adj_R2 = c(summary(mod_mat_struct)$adj.r.squared,
             summary(mod_mat_full)$adj.r.squared,
             summary(mod_por_struct)$adj.r.squared,
             summary(mod_por_full)$adj.r.squared))

comparison |>
  gt(groupname_col = "Subject") |>
  tab_header(title = "Model Comparison",
             subtitle = "Structural vs Full model across both subjects") |>
  fmt_number(columns = c(AIC, RMSE, MAE, CV_RMSE, CV_MAE), decimals = 2) |>
  fmt_number(columns = c(R2, Adj_R2), decimals = 3) |>
  cols_label(Adj_R2 = "Adj. R²", R2 = "R²",
             CV_RMSE = "CV-RMSE", CV_MAE = "CV-MAE") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_row_groups())

```

Across both subjects, the full model outperforms the structural model on
every metric. The AIC drops by around 20 points for Mathematics and 29
points for Portuguese. Thus, the behavioral predictors justify their
inclusion even after accounting for the added complexity.

The clearest way to see the gain is through Adj. R\^2: it roughly
doubles in both subjects, from 0.068 to 0.141 in Mathematics and from
0.094 to 0.187 in Portuguese. This directly answers the core research
question --- personal agency variables do add explanatory power beyond
structural constraints, and the contribution is substantial relative to
what the structural model achieves on its own. Portuguese shows a
slightly larger gain overall, consistent with the stronger and more
consistent signals we observed in the interpretation section.

In terms of actual prediction accuracy, the CV-RMSE provides the more
honest comparison, as it approximates the error each model would make on
an observation it has not seen. The full model improves on the
structural model in both subjects: CV-RMSE drops from 3.24 to 3.15 in
Mathematics and from 2.52 to 2.41 in Portuguese, with CV-MAE showing the
same direction. That said, the absolute prediction errors remain large:
the best-performing model, the full Portuguese model, still carries a
CV-RMSE of 2.41, meaning a typical out-of-sample prediction error of
just under two and a half grade points on a 1--20 scale. So while the
full model is the better model by every criterion, neither model
predicts grades particularly well in absolute terms. An Adj. R\^2 of
0.187 and 0.141 means roughly 80% of the variation in Portuguese grades
--- and nearly 86% in Mathematics --- is left unexplained.

### Model Diagnostics

In this subsections we discuss how reliable are the observations we
made.

```{r}
#| label: diag_helper

make_diag_plots = function(model, title_str) {
  d = augment(model)
  
  p_res = ggplot(d, aes(.fitted, .resid)) +
    geom_point(alpha = 0.3,
               colour = viridis(1, option = VIR_PAL, begin = 0.2)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_smooth(se = FALSE, method = "loess",
                colour = viridis(1, option = VIR_PAL, begin = 0.8)) +
    labs(title = "Residuals vs Fitted",
         x = "Fitted Values", y = "Residuals")

  p_qq = ggplot(d, aes(sample = .std.resid)) +
    stat_qq(alpha = 0.3,
            colour = viridis(1, option = VIR_PAL, begin = 0.2)) +
    stat_qq_line(linetype = "dashed", colour = "grey50") +
    labs(title = "Normal Q-Q",
         x = "Theoretical Quantiles", y = "Standardized Residuals")

  p_sl = ggplot(d, aes(.fitted, sqrt(abs(.std.resid)))) +
    geom_point(alpha = 0.3,
               colour = viridis(1, option = VIR_PAL, begin = 0.2)) +
    geom_smooth(se = FALSE, method = "loess",
                colour = viridis(1, option = VIR_PAL, begin = 0.8)) +
    labs(title = "Scale-Location",
         x = "Fitted Values", y = "sqrt(|Standardized Residuals|)")

  p_lev = ggplot(d, aes(.hat, .std.resid)) +
    geom_point(aes(size = .cooksd), alpha = 0.3,
               colour = viridis(1, option = VIR_PAL, begin = 0.2)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_smooth(se = FALSE, method = "loess",
                colour = viridis(1, option = VIR_PAL, begin = 0.8)) +
    labs(title = "Residuals vs Leverage",
         x = "Leverage", y = "Standardized Residuals", size = "Cook's D")

  (p_res | p_qq) / (p_sl | p_lev) +
    plot_annotation(
      title = title_str,
      theme = theme(plot.title = element_text(size = 14,
                                              face = "bold", hjust = 0.5)))
}

```

#### Mathematics - Full Model

```{r}
#| label: diag_mat_full
#| fig-width: 10
#| fig-height: 8

make_diag_plots(mod_mat_full, "Diagnostics - Mathematics, Full Model")

```

#### Portuguese - Full Model

```{r}
#| label: diag_por_full
#| fig-width: 10
#| fig-height: 8

make_diag_plots(mod_por_full, "Diagnostics - Portuguese, Full Model")
```

The diagnostic plots for both full models look healthy overall, with no
severe violations of the core OLS assumptions. The residuals vs fitted
plots show no obvious non-linearity in either subject, with the loess
line sitting close to zero across the fitted range. The Q-Q plots
confirm approximate normality of residuals in both cases, with only
minor tail deviations in Mathematics. The scale-location plots show some
variance instability in Mathematics suggesting mild heteroscedasticity,
which means confidence intervals may be slightly optimistic and
borderline results such as weekend alcohol should be treated with some
caution. For Portuguese, the main thing to flag is a single
high-leverage point with a noticeably larger Cook's D than the rest of
the sample --- this is almost certainly the student with a final grade
of 1 noted in the IDA. Their influence on the Portuguese estimates is
limited but not negligible. Overall, we think that the models are
reliable enough to support the interpretations made.

### Non-Linear Relationships

For the ordinal predictors encoded with successive difference contrasts
--- `studytime`, `Medu`, `Fedu`, and `traveltime` --- the question of
non-linearity is already addressed by the model specification itself.
Each level transition receives its own coefficient, so the model is not
forced to assume a linear step across categories. The interpretation
sections already highlighted the most notable example of this: the
non-monotone pattern in `studytime`, where the meaningful grade gain is
present only at the 2–5 h to 5–10 h jump.

The variables that were not checked are `health`, `goout`, and `Walc`,
which were treated as strictly linear numeric predictors, and `age`. For
these, we examine the relationship with the final grade graphically.

```{r}
#| label: nonlinear_check
#| fig-width: 14
#| fig-height: 10

d_mat_plot2 = d_mat_analysis |>
  mutate(
    health = factor(health, levels = 1:5,
                    labels = c("1 - very bad", "2", "3", 
                               "4", "5 - very good")),
    goout = factor(goout, levels = 1:5,
                   labels = c("1 - very low", "2", "3", 
                              "4", "5 - very high")),
    Walc = factor(Walc, levels = 1:5,
                  labels = c("1 - very low", "2", "3", 
                             "4", "5 - very high")))

p_health = ggplot(d_mat_plot2, aes(x = health, y = G3)) +
  geom_boxplot(fill = viridis(1, option = VIR_PAL,
                              begin = 0.3), alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 3, color = "white") +
  labs(title = "Health vs G3 (Mathematics)", 
       x = "Health", y = "Final Grade")

p_goout = ggplot(d_mat_plot2, aes(x = goout, y = G3)) +
  geom_boxplot(fill = viridis(1, option = VIR_PAL,
                              begin = 0.5), alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 3, color = "white") +
  labs(title = "Going Out vs G3 (Mathematics)", 
       x = "Going Out", y = "Final Grade")

p_walc = ggplot(d_mat_plot2, aes(x = Walc, y = G3)) +
  geom_boxplot(fill = viridis(1, option = VIR_PAL, 
                              begin = 0.7), alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 3, color = "white") +
  labs(title = "Weekend Alcohol vs G3 (Mathematics)", 
       x = "Weekend Alcohol", y = "Final Grade")

p_age = ggplot(d_mat_analysis, aes(x = age, y = G3)) +
  geom_point(alpha = 0.3, 
             colour = viridis(1, option = VIR_PAL, begin = 0.2)) +
  geom_smooth(method = "loess", span = 0.9, se = FALSE,
              colour = viridis(1, option = VIR_PAL, begin = 0.8)) +
  geom_smooth(method = "lm", se = FALSE, 
              linetype = "dashed", colour = "grey50") +
  labs(title = "Age vs G3 (Mathematics)", 
       x = "Age", y = "Final Grade")

(p_health | p_goout) / (p_walc | p_age) +
  plot_annotation(
    title = "Non-Linearity Check - Numeric Predictors (Mathematics)",
    subtitle = "white diamond marks the group mean; 
    black line marks the median; 
    orange line denote smoothed by loess line",
    theme = theme(plot.title = element_text(size = 14, 
                                            face = "bold", hjust = 0.5),
                  plot.subtitle = element_text(size = 14, 
                                              colour = "grey40", hjust = 0.5)))

```

```{r}
#| label: nonlinear_check_por
#| fig-width: 14
#| fig-height: 10

d_por_plot2 = d_por_analysis |>
  mutate(
    health = factor(health, levels = 1:5,
                    labels = c("1 - very bad", "2", "3", "4", "5 - very good")),
    goout = factor(goout, levels = 1:5,
                   labels = c("1 - very low", "2", "3", "4", "5 - very high")),
    Walc = factor(Walc, levels = 1:5,
                  labels = c("1 - very low", "2", "3", "4", "5 - very high")))

p_health_por = ggplot(d_por_plot2, aes(x = health, y = G3)) +
  geom_boxplot(fill = viridis(1, option = VIR_PAL, 
                              begin = 0.3), alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 3, color = "white") +
  labs(title = "Health vs G3 (Portuguese)", 
       x = "Health", y = "Final Grade")

p_goout_por = ggplot(d_por_plot2, aes(x = goout, y = G3)) +
  geom_boxplot(fill = viridis(1, option = VIR_PAL, 
                              begin = 0.5), alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 3, color = "white") +
  labs(title = "Going Out vs G3 (Portuguese)",
       x = "Going Out", y = "Final Grade")

p_walc_por = ggplot(d_por_plot2, aes(x = Walc, y = G3)) +
  geom_boxplot(fill = viridis(1, option = VIR_PAL, begin = 0.7), alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 3, color = "white") +
  labs(title = "Weekend Alcohol vs G3 (Portuguese)", 
       x = "Weekend Alcohol", y = "Final Grade")

p_age_por = ggplot(d_por_analysis, aes(x = age, y = G3)) +
  geom_point(alpha = 0.3, colour = viridis(1, option = VIR_PAL, begin = 0.2)) +
  geom_smooth(method = "loess", span = 0.9, se = FALSE,
              colour = viridis(1, option = VIR_PAL, begin = 0.8)) +
  geom_smooth(method = "lm", se = FALSE, 
              linetype = "dashed", colour = "grey50") +
  labs(title = "Age vs G3 (Portuguese)", 
       x = "Age (mean-centered)", y = "Final Grade")

(p_health_por | p_goout_por) / (p_walc_por | p_age_por) +
  plot_annotation(
    title = "Non-Linearity Check - Numeric Predictors (Portuguese)",
    subtitle = "white diamond marks the group mean; 
    black line marks the median;
    orange line denote smoothed by loess line",
    theme = theme(plot.title = element_text(size = 14, 
                                            face = "bold", hjust = 0.5),
                  plot.subtitle = element_text(size = 14, 
                                              colour = "grey40", hjust = 0.5)))

```

For **health** in both Mathematics and Portugues, the pattern is
non-monotone --- grades drop at level 3 (good health), and then go up.
The negative linear coefficient in the model is therefore not something
meaningful. For **going out** and **weekend alcohol**, the relationship
seem to be relatively linear (there are deviations, `goout` has a jump
from very low to low; but we think they are negligible given the
significance and results being consistent with the expectations). For
**age**, the loess and linear fits are very close for Mathematics, but
for Portuguese we see a drastic drop at the right tail --- which is
driven by outlier with grade 5.

As a consequence, the most defensible correction would be for health
(and maybe goout) to encode them as ordered factor with successive
difference contrasts, as was done for studytime. However, as we
discussed in the analysis plan this yields 4 (8) more coefficients,
which we were not sure we could afford with this sample size; if we
decided to encode all of the ordinal factors in this way, we would
overwhelm the model --- so we should have treated some of the ordinals
as numeric anyway --- and we did not expect that health has a non-linear
pattern.

But again, looking at it retrospectively, the health should have been
recoded; and it would most likely remove or explain our finding of
negative association with the final grade when treated as just numeric.

## Conclusion

This study looks at whether personal agency --- defined as
student-driven behavioral choices --- helps predict academic success in
Portuguese and Mathematics, beyond structural and environmental factors.
The short answer is yes, but the effect is modest and comes with some
important notes.

In both subjects, the full model performed better than the structural
model on every measure. Adjusted R\^2 roughly doubled, and AIC dropped
by about 20 points. The behavioral patterns were clear: spending more
time **going out** was linked to lower grades, while increasing **study
time** from moderate (2–5 hours) to higher levels (5–10 hours) per week
led to the largest improvement in grades. These patterns were consistent
across subjects and models, which makes them more convincing.

Structural variables, on the other hand, mostly did not show strong
effects. Family background, parental education, travel time, and
internet access had little independent impact. The only small exception
was **parental education**, which showed some effect at higher levels in
Portuguese, but not in Mathematics.

Among the control variables, **sex** showed the most consistent pattern.
It was important in all models and flipped between subjects—male
students did better in Mathematics, while female students did better in
Portuguese. What’s interesting is how this changed after adding
behavioral factors. In Portuguese, the female advantage became smaller,
suggesting that differences in behavior --- like studying more --- help
explain part of the gap. In Mathematics, however, the male advantage
actually became larger after accounting for behavior. This suggests that
these behavioral differences were helping to reduce the gap in this
case.

**Age** showed a different pattern. It had no clear effect in
Mathematics, but in Portuguese it became a positive and significant
factor, but only after behavioral variables were added. This suggests
that behavior was hiding the effect of age in the simpler model. No
similar pattern appeared in Mathematics.

However, the findings must be read critically, and the following
limitations should be kept in mind.

### Limitations

**Low absolute explanatory power.** Even the best-performing model ---
the full Portuguese model --- explains only 18.7% of the variance in
final grades (Adj. R\^2 = 0.187).

**Correlational design.** All inferences are associational.

**Self-report bias.** All behavioral variables (alcohol consumption,
study time, going out, ...) are self-reported.

**Sample scope and generalisability.** The analytic sample consists of
327 students from two schools in Portugal. Both schools are located in
the same region, the sample is heavily urban, and the data are
disproportionately drawn from Gabriel Pereira, with only 32 students
from Mousinho da Silveira.

**Health encoding and the counterintuitive negative association.** The
consistent negative link between self-reported health and grades across
all four models is most likely due to how the variable was coded, rather
than a real effect.

**Small or sparse contrast groups.** Several binary predictors have
severely imbalanced distributions: only 32 students attend Mousinho da
Silveira, only 49 students lack home internet access, and the "Apart"
category for parental status is sparsely populated.
