---
title: "Seminar: Statistics for Data Science: Group C"
author: "Dmytro Tupkalenko (89252101), ..."
date: "May 2026"
date-format: "MMMM YYYY"
format: 
  html:
    toc: true
    toc_float: yes
    toc_depth: 3
    embed-resources: true
    theme: "lux"
execute: 
  warning: false
  messages: false
  echo: true
  error: false
---



## Setup

```{r}
#| label: setup


# Packages used
library(tidyverse)
library(naniar)  
library(Hmisc)     
library(kableExtra)
library(GGally)   
library(gt)
library(reshape2)
library(htmltools)
library(patchwork)
library(psych)
library(viridis)

# ── Global plot settings ──────────────────────────────────────────────────────
# Viridis palette used throughout: "A"=magma  "B"=inferno  "C"=plasma
#                                  "D"=viridis "E"=cividis  "H"=turbo
VIR_PAL = "B"
theme_set(theme_minimal(base_size = 11))
setwd("./")

```



## Data Management

**Read datasets**
```{r}
#| label: data_import

d_mat = read.table("data/student-mat.csv", sep=";", header=TRUE)
d_por = read.table("data/student-por.csv", sep=";", header=TRUE)

```

**We will work on both Mathematics and Portuguese - so merge them**
```{r}
#| label: data_merge

# NOTE: we use bigger set of variables than in the description of the seminar,
# because we saw that in the case of using the smaller set, we get 
# non-consistent rows, so two rows in mat and por with the same signature
# gave different responses to e.g. activities
d_combined = merge(d_mat, d_por, 
           by = c("school", "sex", "age", "address", "famsize", "Pstatus", 
                  "Medu", "Fedu", "Mjob", "Fjob", "reason", 
                  "guardian", "traveltime", "studytime", "activities", 
                  "nursery", "higher", "internet", "romantic", 
                  "famrel", "freetime", "goout", "Dalc", "Walc", "health"),
           suffixes = c(".mat", ".por"))

cat("Number of overlapping students:", nrow(d_combined), "\n")

```

**Recode Variables**
```{r}
#| label: data_transform

#| label: data_transform

# RECODE KEY VARIABLES (Shared across both datasets)
# These do NOT have suffixes because they were used to merge.

# Binary Factors
binary_map = list(
  school   = list(lv = c("GP", "MS"), 
                  lb = c("0 - Gabriel Pereira", "1 - Mousinho da Silveira")),
  sex      = list(lv = c("F", "M"),   
                  lb = c("0 - Female", "1 - Male")),
  address  = list(lv = c("U", "R"),   
                  lb = c("0 - Urban", "1 - Rural")),
  famsize  = list(lv = c("LE3", "GT3"), 
                  lb = c("0 - <=3 members", "1 - >3 members")),
  Pstatus  = list(lv = c("A", "T"),   
                  lb = c("0 - Apart", "1 - Living together")),
  nursery  = list(lv = c("no", "yes"), 
                  lb = c("0 - No", "1 - Yes")),
  internet = list(lv = c("no", "yes"), 
                  lb = c("0 - No", "1 - Yes")),
  higher   = list(lv = c("no", "yes"), lb = c("0 - No", "1 - Yes")),
  activities = list(lv = c("no", "yes"), 
                    lb = c("0 - No", "1 - Yes")),
  romantic   = list(lv = c("no", "yes"), 
                    lb = c("0 - No", "1 - Yes"))
)

for (v in names(binary_map)) {
  d_combined[[v]] = factor(d_combined[[v]], 
                           levels = binary_map[[v]]$lv, 
                           labels = binary_map[[v]]$lb)
}

# Nominal Factors
d_combined$Mjob = factor(d_combined$Mjob, 
                         levels = c("at_home", "health", "other", 
                                    "services", "teacher"),
                         labels = c("0 - At home", "1 - Health", 
                                    "2 - Other", "3 - Services", "4 - Teacher"))
d_combined$Fjob = factor(d_combined$Fjob, 
                         levels = c("at_home", "health", "other", 
                                    "services", "teacher"),
                         labels = c("0 - At home", "1 - Health", 
                                    "2 - Other", "3 - Services", "4 - Teacher"))
d_combined$reason = factor(d_combined$reason, 
                           levels = c("course", "home", "other", "reputation"),
                           labels = c("0 - Course preference", 
                                      "1 - Close to home", "2 - Other", 
                                      "3 - School reputation"))
d_combined$guardian = factor(d_combined$guardian, 
                             levels = c("mother", "father", "other"),
                             labels = c("0 - Mother", 
                                        "1 - Father", "2 - Other"))

# Ordered Factors (Shared)
edu_levels = c("0 - None", "1 - Primary (4th grade)", "2 - 5th-9th grade", 
               "3 - Secondary", "4 - Higher education")
d_combined$Medu = factor(d_combined$Medu, 
                         levels = 0:4, 
                         labels = edu_levels, 
                         ordered = TRUE)
d_combined$Fedu = factor(d_combined$Fedu, 
                         levels = 0:4, 
                         labels = edu_levels, 
                         ordered = TRUE)

d_combined$traveltime = factor(d_combined$traveltime, 
                               levels = 1:4, 
                               labels = c("1 - <15 min", "2 - 15-30 min", 
                                          "3 - 30 min-1 h", "4 - >1 h"), 
                               ordered = TRUE)
d_combined$studytime = factor(d_combined$studytime, 
                              levels = 1:4, 
                              labels = c("1 - <2 h", "2 - 2-5 h", 
                                         "3 - 5-10 h", "4 - >10 h"), 
                              ordered = TRUE)

# Shared Likert Scales (1-5)
likert_vars = list(
  famrel = c("very bad", "excellent"),
  freetime = c("very low", "very high"),
  goout = c("very low", "very high"),
  Dalc = c("very low", "very high"),
  Walc = c("very low", "very high"),
  health = c("very bad", "very good")
)

for (v in names(likert_vars)) {
  d_combined[[v]] = factor(d_combined[[v]], 
                           levels = 1:5, 
                           ordered = TRUE,
                           labels = 
                             c(paste0("1 - ", likert_vars[[v]][1]), "2", 
                                      "3", "4", 
                               paste0("5 - ", likert_vars[[v]][2])))
}

# RECODE SUBJECT-SPECIFIC VARIABLES
# These DO have suffixes because they differed between files
for (sfx in c(".mat", ".por")) {
  
  # Failures
  d_combined[[paste0("failures", sfx)]] = 
    factor(d_combined[[paste0("failures", sfx)]],
           levels = 0:3, 
           ordered = TRUE,
           labels = c("0 - 0", "1 - 1", "2 - 2", "3 - 3+"))
  
  for (v in c("schoolsup", "famsup", "paid")) {
    d_combined[[paste0(v, sfx)]] = factor(d_combined[[paste0(v, sfx)]],
                                         levels = c("no", "yes"),
                                         labels = c("0 - No", "1 - Yes"))
  }
}

```



## Research Question - General Formulation

**"Personal Agency vs. Environmental Determinism: To what extent do student-driven behavioral choices predict academic success compared to structural and environmental constraints?"**

**Variable selection**
```{r}
#| label: analysis_variable_selection

# General Variables
general_vars = c("school", "sex", "age")

# Behavioral: Choices and habits 
behavioural_vars = c("goout", "activities", "Dalc", "Walc", 
                     "studytime", "romantic")

# Structural: Background and constraints 
structural_vars = c("address", "famsize", "Pstatus", "Medu", 
                    "Fedu", "traveltime", "health")


# Function to create subject-specific data
prepare_subject_data = function(df, subject_suffix) {
  # Shared columns do not have suffixes because they were used as keys in merge
  shared_cols = c(general_vars, behavioural_vars, structural_vars)
  
  # Only G3 (the target) has suffix
  target_cols = c(shared_cols, paste0("G3", subject_suffix))
  
  df_sub = df[, target_cols]
  colnames(df_sub) = gsub(subject_suffix, "", colnames(df_sub), fixed = TRUE)
  return(df_sub)
}

# Create more specific datasets
d_mat_analysis = prepare_subject_data(d_combined, ".mat")
d_por_analysis = prepare_subject_data(d_combined, ".por")

```



## Initial Data Analysis

### IDA Screening: Missing values
```{r}
#| label: ida_missing_values

# All predictor variables are shared (merge keys) - only G3 differs by subject
# Build a single unified view: shared vars once + G3 for each subject
shared_vars_ida = setdiff(names(d_mat_analysis), "G3")
d_unified_ida = d_mat_analysis[, shared_vars_ida]
d_unified_ida$G3_Math = d_mat_analysis$G3
d_unified_ida$G3_Por = d_por_analysis$G3

d_unified_ida |>
  miss_var_summary() |>
  gt() |>
  tab_header(
    title    = "Missing Value Summary",
    subtitle = "Predictor variables are shared across subjects; 
    G3 shown separately per subject")

```

**Zero-values - These are valid encoded values but worth flagging as potential edge cases**
```{r}
#| label: ida_zero_check

# Medu / Fedu are shared (merge keys) -> one count suffices
# G3 may differ by subject -> shown separately
zero_summary = data.frame(
  Variable = c("Medu = None", "Fedu = None", "G3 = 0 (Math)", 
               "G3 = 0 (Portuguese)"),
  N = c(sum(d_mat_analysis$Medu == "0 - None"),
        sum(d_mat_analysis$Fedu == "0 - None"),
        sum(d_mat_analysis$G3 == 0),
        sum(d_por_analysis$G3 == 0)))

zero_summary |>
  gt() |>
  tab_header(
    title    = "Zero / Baseline Value Counts",
    subtitle = "Medu & Fedu: parents with no education; 
    G3: students scoring 0") |>
  cols_label(Variable = "Variable", N = "n")

```


### IDA Screening: Univariate descriptions

**Categorical variables**
```{r}
#| label: ida_univariate_categorical
#| fig-width: 10
#| fig-height: 30

cat_vars = c("school", "sex", "address", "famsize", "Pstatus",
             "Medu", "Fedu", "traveltime", "studytime",
             "activities", "romantic",
             "goout", "Dalc", "Walc", "health")

subj_cols = viridis(2, option = VIR_PAL, begin = 0.1, end = 0.5)

make_cat_plot = function(df, var, fill_col) {
  ggplot(df, aes(x = .data[[var]])) +
    geom_bar(fill = fill_col, color = "white", width = 0.7) +
    geom_text(stat = "count", aes(label = after_stat(count)),
              vjust = -0.5, size = 4, color = "grey20") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          panel.grid.major.x = element_blank(),
          plot.title= element_text(size = 12, face = "bold")) +
    labs(title = var, x = NULL, y = NULL)
}

plots_shared = lapply(cat_vars, make_cat_plot,
                      df = d_mat_analysis, fill_col = subj_cols[1])

wrap_plots(plots_shared, ncol = 2) +
  plot_annotation(
    title = "Univariate Distributions - Categorical Variables",
    subtitle = "Shared across Math & Portuguese (overlapping students only)",
    theme = theme(plot.title = element_text(size = 16, face = "bold", 
                                                  hjust = 0.5),
                     plot.subtitle = element_text(size = 12, hjust = 0.5)))

```

**Continuous variables**
```{r}
#| label: ida_univariate_continuous
#| fig-width: 12
#| fig-height: 10

make_hist = function(df, var, fill_col, subject, binw = 1) {
  ggplot(df, aes(x = .data[[var]])) +
    geom_histogram(binwidth = binw, fill = fill_col,
                   color = "white", alpha = 0.9) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          axis.text = element_text(size = 10),
          plot.title = element_text(size = 12, face = "bold")) +
    labs(title = paste0(var, " - ", subject), x = var, y = NULL)
}

p_age = make_hist(d_mat_analysis, "age", subj_cols[1], "Shared",
                     binw = 1)
p_g3_mat = make_hist(d_mat_analysis, "G3", subj_cols[1], "Math", 
                     binw = 1)
p_g3_por = make_hist(d_por_analysis, "G3", subj_cols[2], "Portuguese", 
                     binw = 1)

(p_age / (p_g3_mat | p_g3_por)) +
  plot_annotation(
    title = "Univariate Distributions — Continuous Variables",
    theme = theme(plot.title = element_text(size = 16, face = "bold", 
                                            hjust = 0.5)))

```


```{r}
#| label: ida_continuous_summary

make_desc_table = function(df, title) {
  vars <- names(df)
  psych::describe(df) |>
    select(mean, sd, median, min, max) |>
    mutate(IQR = sapply(vars, function(v) IQR(df[[v]], na.rm = TRUE))) |>
    select(mean, sd, median, IQR, min, max) |>
    as.data.frame() |>
    rownames_to_column("variable") |>
    gt() |>
    tab_header(title = title) |>
    fmt_number(columns = where(is.double), decimals = 2) |>
    as_raw_html()
}

# Age is shared — describe once.
# G3 differs by subject — include both in a single table.
cont_vars_unified = data.frame(
  age = d_mat_analysis$age,       # identical in both datasets
  G3_Mat = d_mat_analysis$G3,
  G3_Por = d_por_analysis$G3
)

HTML(make_desc_table(cont_vars_unified,
                     "Continuous Variables - Shared (age) & Per Subject (G3)"))

```



